---
title: "tidyverse_with_gapminder"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# tidyverse R Markdown

This is an R Markdown file capturing the exercises used in a datacamp tidyverse introductory course. The dataset was the gapminder data found in the tidyverse package.

## Import the libraries and data
```{r}
library(gapminder)
library(dplyr)
gapminder
```

## Filter Verb

Filters show a subset of the data

Filter for just the year 2007 using the pipe %>%
Piping simply is creating a dataframe on the fly without
changing your original data frame unless you assign it

```{r}
gapminder %>%
  filter(year == 2007)

# filter for the US - need quotes around text
gapminder %>%
  filter(country == "United States")

# filter by two conditions
gapminder %>%
  filter(country == "United States", year == 2007)
```

## Arrange verb

Arrange verb sorts the data frame.

```{r}
gapminder %>%
  arrange(gdpPercap)

gapminder %>%
  arrange(desc(gdpPercap))

# combining verbs to find richest countries in 2007 descending order
gapminder %>%
  filter(year == 2007) %>%
  arrange(desc(gdpPercap))
```

## Mutate verb

Mutate allows you to change the data frame without changing the original data.

```{r}
gapminder %>%
  mutate(pop = pop / 1000000)

# add a variable
gapminder %>%
  mutate(gdp = gdpPercap * pop) %>%
  arrange(desc(gdp))

# Create, filter, and sort the data by combining verbs
gapminder %>%
  mutate(gdp = gdpPercap * pop) %>%
  filter(year == 2007) %>%
  arrange(desc(gdp))
```

## Data Visualization

Use ggplot for creating a viz. 
Rules: 1) data, 2) aesthetic and 3) layer

```{r}
library(ggplot2)

# create a new dataframe for 2007
gapminder_2007 <- gapminder %>%
  filter(year == 2007)

gapminder_2007

# ggplots start with the df and then an aesthetic
# then you add a layer (scatterplot, bar chart, etc.)
ggplot(gapminder_2007, aes(x = gdpPercap,
                           y = lifeExp)) +
  geom_point()

# use a logscale if there's several orders of magnitude in one section
# of the x-axis or y-axis
ggplot(gapminder_2007, aes(x = gdpPercap,
                           y = lifeExp)) +
  geom_point() +
  scale_x_log10()

# group the data by continent with color
ggplot(gapminder_2007, aes(x = gdpPercap,
                           y = lifeExp, color = continent)) +
  geom_point() +
  scale_x_log10()

# group the data by continent and size of population as aesthetics #3 and #4
# on top of aesthetics #1 (x) and #2 (y)
ggplot(gapminder_2007, aes(x = gdpPercap,
                           y = lifeExp, 
                           color = continent,
                           size = pop)) +
  geom_point() +
  scale_x_log10()
```

## Faceting

Faceting explores categorical variables by using a series of subplots.

```{r}
ggplot(gapminder_2007, aes(x = gdpPercap,
                           y = lifeExp)) +
  geom_point() +
  scale_x_log10() +
  facet_wrap(~ continent)
```

### Summarize

Use the summarize functionality for more in depth analysis. You can find
powerful data points using summarize and its functionality from the dplyr package. Some functions include:

mean, 
min, 
max, 
sum 
median

You can do as many of these functions as you wish in combination.

```{r}
gapminder %>%
  summarize(meanLifeExp = mean(lifeExp))

gapminder %>%
  filter(year == 2007) %>%
  summarize(meanLifeExp = mean(lifeExp))

gapminder %>%
  filter(year == 2007) %>%
  summarize(meanLifeExp = mean(lifeExp),
            totalPop = sum(pop))
```

### group_by

Use group_by to visualize a "class" of data points within a body of data.

```{r}
gapminder %>%
  group_by(year) %>%
  summarize(meanLifeExp = mean(lifeExp),
            totalPop = sum(pop))

# group_by is now done after filtering and does the summarize by a "group"
gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarize(meanLifeExp = mean(lifeExp),
            totalPop = sum(pop))

# group_by is now done after filtering and does the summarize by a "group"
gapminder %>%
  group_by(year, continent) %>%
  summarize(meanLifeExp = mean(lifeExp),
            totalPop = sum(pop))
```

### dynamic viz

```{r}
by_year <- gapminder %>%
  group_by(year) %>%
  summarize(totalPop = sum(pop),
            meanLifeExp = mean(lifeExp))

# plot these summaries now
ggplot(by_year, aes(x = year, y = totalPop)) +
  geom_point()

# add y = 0 to the graph
ggplot(by_year, aes(x = year, y = totalPop)) +
  geom_point() +
  expand_limits(y = 0)

# group by year, continent into new data frame
by_year_continent <- gapminder %>%
  group_by(year, continent) %>%
  summarize(totalPop = sum(pop),
            meanLifeExp = mean(lifeExp))

ggplot(by_year_continent, aes(x = year,
                              y = totalPop,
                              color = continent)) +
  geom_point() +
  expand_limits(y = 0)
```

### Line plots

Used for time series visualization, somewhat easier than plotting points.

```{r}
ggplot(by_year_continent, aes(x = year,
                              y = meanLifeExp,
                              color = continent)) +
  geom_line() +
  expand_limits(y = 0)
```

### Bar plots

Used to display heights (y-axis) broken down by a categorical variable (x-axis).

Use this for comparative analysis.

```{r}
by_continent <- gapminder %>%
  group_by(continent) %>%
  summarize(totalPop = sum(pop),
            meanLifeExp = mean(lifeExp))

ggplot(by_continent, aes(x = continent,
                         y = meanLifeExp)) +
  geom_col()
```

### Histograms

Used to look at one variable closely (on the x-axis).

Uses "bins". You can change the bin width, the # of observations in a bin, etc.

```{r}
ggplot(gapminder_2007, aes(x = lifeExp)) +
  geom_histogram()

# change the histogram bin width
# you can also change the # of bins, using bins = <insert #>
ggplot(gapminder_2007, aes(x = lifeExp)) +
  geom_histogram(binwidth = 5)
```

### Box plots

Used to 1) understand the distribution of a variable and 2) compare across a set of predictors against a response variable.

Breaks down into quadrants (50% shown in the "box"), shows the median (dark line) and stretches with whiskers (3rd and 4th quadrants).

```{r}
ggplot(gapminder_2007, aes(x = continent, y = lifeExp)) +
  geom_boxplot()

# add a label using labs and title
ggplot(gapminder_2007, aes(x = lifeExp)) +
  geom_histogram(binwidth = 5) +
  labs(
    title = paste(
      "Life Expectancy Across Continents"
    )
  )

# done another way
ggplot(gapminder_2007, aes(x = lifeExp)) +
  geom_histogram(binwidth = 5) +
  labs(title = "Life Expectancy Across Continents")

# using ggtitle
ggplot(gapminder_2007, aes(x = lifeExp)) +
  geom_histogram(binwidth = 5) +
  ggtitle("Life Expectancy Across Continents")
```


